---
title: "PAES_heatmap"
author: "Victoria Orr"
date: "2025-11-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

```
Data analysis for manuscript "The pQBR mercury resistance plasmids: a model set of sympatric environmental mobile genetic elements". Code written by Victoria T Orr (Victoria.Orr@liverpool.ac.uk), and James P.J. Hall. 


First get mash distances for plasmids:
2_pP19E3.1.fna
4_pJBCL41.fna
6_pBT2436.fna
9_pQBR103.fna
10_pQBR57.fna


```{bash}

mash sketch -o plasmids *.fna
mash dist plasmids.msh plasmids.msh > mash_distances.tab
```

```{r}
mash <- read.table("mash_distances.tab", header = FALSE, sep = "\t")
colnames(mash) <- c("query", "reference", "distance", "p_value", "shared_hashes")

plasmid_names <- c(
  "10_pQBR57.fna" = "pQBR57 (IV)",
  "9_pQBR103.fna" = "pQBR103 (I)",
  "4_pJBCL41.fna" = "pJBCL41",
  "2_pP19E3.1.fna" = "pP19E3.1",
  "6_pBT2436.fna" = "pBT2436"
)

mash$query <- plasmid_names[mash$query]
mash$reference <- plasmid_names[mash$reference]

plasmids <- unique(c(mash$query, mash$reference))
dist_mat <- matrix(0, nrow = length(plasmids), ncol = length(plasmids),
                   dimnames = list(plasmids, plasmids))

for (i in seq_len(nrow(mash))) {
  q <- mash$query[i]
  r <- mash$reference[i]
  d <- mash$distance[i]
  dist_mat[q, r] <- d
  dist_mat[r, q] <- d
}
dist_obj <- as.dist(dist_mat)
hc <- hclust(dist_obj, method = "average")
plasmid_order <- hc$labels[hc$order]

mash_long <- as.data.frame(as.table(dist_mat)) %>%
  rename(query = Var1, reference = Var2, distance = Freq)


ggplot(mash_long, aes(
  x = factor(reference, levels = plasmid_order),
  y = factor(query, levels = rev(plasmid_order)),   # reverse for top-down match
  fill = distance
)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "black", name = "MASH Distance") +
  labs(x = "Plasmid", y = "Plasmid") +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10)
  )

ggsave("PAESheatmap.png", dpi=500, height = 5, width = 6)