---
title: "Creating tree for Group III"
output: html_document
date: "2025-10-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Data analysis for manuscript "The pQBR mercury resistance plasmids: a model set of sympatric environmental mobile genetic elements". Code written by Victoria T Orr (Victoria.Orr@liverpool.ac.uk), and James P.J. Hall. 

Code for creating trees and heatmaps for Group III plasmids.
```{r}
library(tidyverse)
library(ggtree)
library(ape)
library(phangorn)
library(ggtree)
library(tidytree)
library(patchwork)

```

```{r}
group_iii_rel_pirate <- read.table("PIRATE.gene_families.ordered.tsv",
                                 header=TRUE, sep="\t")

giprel_matrix <- data.frame(cbind(group_iii_rel_pirate[2], ifelse(group_iii_rel_pirate[23:length(group_iii_rel_pirate)]=="", 0, 1))) %>%
  column_to_rownames("gene_family")

giprel_genefam_dendro <- hclust(dist(giprel_matrix))  
giprel_plasmid_dendro <- hclust(dist(t(giprel_matrix)))

giprel_genefam_order <- rownames(giprel_matrix)[giprel_genefam_dendro$order]
giprel_plasmid_order <- colnames(giprel_matrix)[giprel_plasmid_dendro$order]

giprel_long <- group_iii_rel_pirate %>% select(gene_family, threshold, starts_with("pQ")) %>%
  pivot_longer(cols = starts_with("pQ"), names_to = "plasmid", values_to = "gene") %>% 
  filter(gene != "") %>%
  mutate(gene_family = factor(gene_family, levels=rev(giprel_genefam_order)),
         plasmid = factor(plasmid, levels=giprel_plasmid_order))

ggplot(giprel_long, aes(x=gene_family, y=plasmid)) + geom_tile()
```

#### Create a core genome tree for Group III and relative

Check alignment from PIRATE. Trim gappy/misaligned columns. Align using IQtree3.

```
-m MFP               Extended model selection followed by tree inference
-B 1000              1000 bootstrap replicates
--alrt 1000          1000 replicates for SH approximate likelihood ratio test
-T AUTO              Automatic thread assignment
```

```{bash, eval=FALSE}
trimal -in core_alignment.fasta \
-out core_rel_aln.trimmed.fasta \
-gt 0.9

iqtree3 -s core_rel_aln.trimmed.fasta  \
-m MFP \
-B 1000 --alrt 1000 \
-T AUTO \
-pre group_i_tree

mv group_i_tree* ./trees/
```

```{r}
if (!exists("is.waive")) {
  is.waive <- \(x) inherits(x, "waiver") # nolint
}
```

```{r}
gi_tr <- read.tree("./trees/group_i_tree.treefile")

tr_mid <- midpoint(gi_tr)

tr_mid$tip.label <- recode(tr_mid$tip.label,
                           "pQG3rel01" = "unnamed1",
                           "pQG3rel02" = "unnamed2",
                           "pQG3rel03" = "unnamed3",
                           "pQG3rel04" = "unnamed4",
                           "pQG3rel05" = "unnamed5",
                           "pQG3rel06" = "unnamed6",
                           )
tb <- as_tibble(tr_mid)

supports <- tb %>%
  transmute(node,
            sh_alrt = as.numeric(str_extract(label, "^[^/]+")),
            ufboot  = as.numeric(str_extract(label, "(?<=/).+$")))

p_sup <- ggtree(tr_mid) +
  theme_tree2() +
  geom_text2(
    aes(subset = !isTip, label =
          ifelse(!is.na(supports$ufboot),
                paste0("SH=", round(supports$sh_alrt), " / UF=", round(supports$ufboot)),
                 round(as.numeric(label)))),
    hjust = -0.1, size = 2.8, color = "grey20"
  ) +
  # xlim(0, max(node.height(tr_mid)) * 1.15) +
  ggtitle("Tree with SH-aLRT / UFBoot support")
p_sup

tree <- ggtree(tr_mid) + geom_tiplab(size=3, align=TRUE) +xlim(0,0.2)

tree_order <- get_taxa_name(p_sup)


giprel_long_filtered$plasmid <- recode(giprel_long$plasmid,
                           "pQG3rel01" = "unnamed1",
                           "pQG3rel02" = "unnamed2",
                           "pQG3rel03" = "unnamed3",
                           "pQG3rel04" = "unnamed4",
                           "pQG3rel05" = "unnamed5",
                           "pQG3rel06" = "unnamed6",
                           )

heatmap <- ggplot(giprel_long_filtered,
                  aes(x=gene_family,y=factor(plasmid,
                                             levels=rev(tree_order)))) +
  geom_tile() +
  ylab(NULL) +
  xlab("Gene Family") +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_text(size=15),
        axis.text.y = element_blank())

tree + heatmap + plot_layout(widths = c(0.9,1))

ggsave("GroupIII_heattree.png", dpi = 500, height=5, width=8)
```
