---
title: "Attempted classification of pQBR plasmids"
date: "revised 2025"
output: rmarkdown::github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list=ls())

library(tidyverse)
library(ggplot2)
library(knitr)

source("./func/theme_pub.r")
theme_github <- function(x){
  theme_pub(base_size=12)
}

theme_set(theme_github())
```


## Classification

Attempt to classify the plasmids according to the IncP scheme outlined by [Nishimura et al. 2024](https://doi.org/10.1101/2024.09.03.610885)

Download sequences described in the paper.

```{r}
incp_groups <- read.csv("./ref/Nishimura2024.csv", header=TRUE, sep=",")

incp_groups %>% kable()
```

Download the sequences. 

```{bash, eval=FALSE}
mkdir -p 3_incp_groups/seqs

tail -n+2  ./ref/Nishimura2024.csv \
  | awk -v FS="," -v OFS="\t" '{print $4}' | while read ACCESSION
  do
  curl "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/efetch.fcgi?db=nucleotide&amp;id=${ACCESSION}&amp;rettype=fasta" > ./3_incp_groups/seqs/${ACCESSION}.fasta
  done
  
cat ./3_incp_groups/seqs/*.fasta > ./3_incp_groups/seqs.fasta
```

Do a MASH comparison.

```{bash, eval=FALSE}
mkdir ./3_incp_groups/mash_results

mash sketch -i -s 100000 ./3_incp_groups/seqs.fasta -o ./3_incp_groups/seqs.msh

cat ./1_sketches/annotated_plasmids.list | while read PLASMID
do
mash dist -p 64 \
  ./1_sketches/${PLASMID}.msh \
  ./3_incp_groups/seqs.msh \
  | sort -gk3 \
  > ./3_incp_groups/mash_results/${PLASMID}_incpref_mash_results.txt
head -n1 ./3_incp_groups/mash_results/${PLASMID}_incpref_mash_results.txt >> ./3_incp_groups/mash_tophits.tsv
done
```

Examine in R.

```{r}
mash_tophits <- read.table("./3_incp_groups/mash_tophits.tsv", header=FALSE,
           col.names=c("query","subject","dist","e_val","num_hits")) %>%
  mutate(plasmid = gsub(".*(pQBR[0-9Rpd]+).fna", "\\1", query),
         Accession = gsub("\\.[0-9]", "", subject),
         N_hits = gsub("/100000", "", num_hits))

mash_tophits %>% left_join(incp_groups, by = "Accession") %>% 
  select(plasmid, dist, N_hits, Inc_group, Alt_group, Representative, Accession, MOB, MPF) %>% 
  arrange(-dist) %>% kable()
```

This table indicates:

- The Group I plasmids match the pQBR103 IncP group defined by Nishimura et al. (pQBR103), as expected. 
- The Group II plasmids (roughly) match the IncPSTY group defined by Nishimura et al. (pSTY)
- The Group IIb plasmid pQBR26 roughly matches the IncP-7 group (pCAR1)
- The plasmids matching the IncP-9 plasmid (pWW0) are likely matching through the Tn4652 transposon, since they include representatives of Group III and Group IV that have Tn4652, and pWW0 has the complete version of this *tol* transposon. 
- The Group III plasmids and Group IV plasmids do not match one of the defined IncP groups. 

---

**[Back to index.](../README.md)**
