---
title: "Attempted classification of pQBR plasmids"
output: rmarkdown::github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list=ls())

library(tidyverse)
library(ggplot2)
library(knitr)

source("./func/theme_pub.r")
theme_github <- function(x){
  theme_pub(base_size=12)
}

theme_set(theme_github())
```


## PInc classification

Attempt to classify the plasmids according to the IncP scheme outlined by [Nishimura et al. 2024](https://doi.org/10.1101/2024.09.03.610885)

Download sequences described in the paper.

```{r}
incp_groups <- read.csv("./ref/Nishimura2024.csv", header=TRUE, sep=",")

incp_groups %>% kable()
```

Download the sequences. 

```{bash, eval=FALSE}
mkdir -p 3_incp_groups/seqs

tail -n+2  ./ref/Nishimura2024.csv \
  | awk -v FS="," -v OFS="\t" '{print $4}' | while read ACCESSION
  do
  curl "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/efetch.fcgi?db=nucleotide&amp;id=${ACCESSION}&amp;rettype=fasta" > ./3_incp_groups/seqs/${ACCESSION}.fasta
  done
  
cat ./3_incp_groups/seqs/*.fasta > ./3_incp_groups/seqs.fasta
```

Do a MASH comparison.

```{bash, eval=FALSE}
mkdir ./3_incp_groups/mash_results

mash sketch -i -s 100000 ./3_incp_groups/seqs.fasta -o ./3_incp_groups/seqs.msh

cat ./1_sketches/annotated_plasmids.list | while read PLASMID
do
mash dist -p 64 \
  ./1_sketches/${PLASMID}.msh \
  ./3_incp_groups/seqs.msh \
  | sort -gk3 \
  > ./3_incp_groups/mash_results/${PLASMID}_incpref_mash_results.txt
head -n1 ./3_incp_groups/mash_results/${PLASMID}_incpref_mash_results.txt >> ./3_incp_groups/mash_tophits.tsv
done
```

Examine in R.

```{r}
mash_tophits <- read.table("./3_incp_groups/mash_tophits.tsv", header=FALSE,
           col.names=c("query","subject","dist","e_val","num_hits")) %>%
  mutate(plasmid = gsub(".*(pQBR[0-9Rpd]+).fna", "\\1", query),
         Accession = gsub("\\.[0-9]", "", subject),
         N_hits = gsub("/100000", "", num_hits))

mash_tophits %>% left_join(incp_groups, by = "Accession") %>% 
  select(plasmid, dist, N_hits, Inc_group, Alt_group, Representative, Accession, MOB, MPF) %>% 
  arrange(-dist) %>% kable()
```

This table indicates:

- The Group I plasmids match the pQBR103 IncP group defined by Nishimura et al. (pQBR103), as expected. 
- The Group II plasmids (roughly) match the IncPSTY group defined by Nishimura et al. (pSTY)
- The Group IIb plasmid pQBR26 roughly matches the IncP-7 group (pCAR1)
- The non-determined plasmid pQBR105 matches IncP-12 group (R716)
- The plasmids matching the IncP-9 plasmid (pWW0) are likely matching through the Tn4652 transposon, since they include representatives of Group III and Group IV that have Tn4652, and pWW0 has the complete version of this *tol* transposon. 
- The Group III plasmids and Group IV plasmids do not match one of the defined IncP groups. 

Run a BLAST comparison to visualise.

```{bash, eval=FALSE}
makeblastdb -dbtype nucl -in ./3_incp_groups/seqs/NC_022739.fasta
makeblastdb -dbtype nucl -in ./3_incp_groups/seqs/AB088420.fasta
makeblastdb -dbtype nucl -in ./3_incp_groups/seqs/LC685026.fasta


mkdir ./3_incp_groups/blast

tblastx -query ./bakta_annotated/pQBR23/pQBR23.fna -db ./3_incp_groups/seqs/NC_022739.fasta \
  -outfmt 6 > ./3_incp_groups/blast/pQBR23_pSTY.tblastx
  
tblastx -query ./bakta_annotated/pQBR26/pQBR26.fna -db ./3_incp_groups/seqs/AB088420.fasta \
  -outfmt 6 > ./3_incp_groups/blast/pQBR26_pCAR1.tblastx
  
tblastx -query ./bakta_annotated/pQBR105/pQBR105.fna -db ./3_incp_groups/seqs/LC685026.fasta \
  -outfmt 6 > ./3_incp_groups/blast/pQBR105_R716.tblastx

```

Examined using ACT. Rough plots here with ggplot.

```{r}
blast_columns <- c("queryId","subjectId","percIdentity", "alnLength",
                   "mismatchCount", "gapOpenCount", "queryStart", "queryEnd", 
                   "subjectStart", "subjectEnd", "eVal", "bitScore")

pQBR23_pSTY_tblastx <- read.table("./3_incp_groups/blast/pQBR23_pSTY.tblastx",
                                  col.names=blast_columns)

makeComparison <- function(x){
  vals <- x %>% mutate(id = row_number()) %>%
    filter(eVal < 1e-100 & alnLength > 100) %>%
    select(id, percIdentity, queryStart, queryEnd, subjectEnd, subjectStart) %>%
    mutate(direction = case_when(queryEnd > queryStart & subjectEnd > subjectStart ~ "F",
                                 queryEnd < queryStart & subjectEnd > subjectStart ~ "R",
                                 queryEnd > queryStart & subjectEnd < subjectStart ~ "R",
                                 queryEnd < queryStart & subjectEnd < subjectStart ~ "F")) %>%
    pivot_longer(cols = c("queryStart","queryEnd","subjectEnd","subjectStart"),
                 names_to = "type",
                 values_to = "x_coord")
  
  vals <- vals %>%
    mutate(y_coord = rep(c(1,1,2,2), length.out = nrow(vals)))
  
  ggplot(data = vals) +
    geom_polygon(aes(x = x_coord, y=(0-y_coord), group=id, fill = direction, alpha=percIdentity))
}

(pQBR23_pSTY_comparisons <- makeComparison(pQBR23_pSTY_tblastx))

pQBR26_pCAR1_tblastx <- read.table("./3_incp_groups/blast/pQBR26_pCAR1.tblastx",
                                  col.names=blast_columns)
(pQBR26_pCAR1_comparisons <- makeComparison(pQBR26_pCAR1_tblastx))

pQBR105_R716_tblastx <- read.table("./3_incp_groups/blast/pQBR105_R716.tblastx",
                                  col.names=blast_columns)
(pQBR105_R716_comparisons <- makeComparison(pQBR105_R716_tblastx))
```

pQBR23 and pQBR26 match their corresponding groups well, but pQBR105 matches its corresponding plasmid R716 only through a transposon.

```{bash, eval=FALSE}
blastn -query ./bakta_annotated/pQBR23/pQBR23.fna -db ./3_incp_groups/seqs/NC_022739.fasta \
  -perc_identity 80 \
  -outfmt "6 std qcovus" # 56% coverage at >80% identity
  
blastn -query ./bakta_annotated/pQBR24/pQBR24.fna -db ./3_incp_groups/seqs/NC_022739.fasta \
  -perc_identity 80 \
  -outfmt "6 std qcovus" # 56% coverage at >80% identity

blastn -query ./bakta_annotated/pQBR26/pQBR26.fna -db ./3_incp_groups/seqs/AB088420.fasta \
  -perc_identity 80 \
  -outfmt "6 std qcovus" # 42% coverage at >80% identity

blastn -query ./bakta_annotated/pQBR105/pQBR105.fna -db ./3_incp_groups/seqs/LC685026.fasta  \
  -perc_identity 80 \
  -outfmt "6 std qcovus" # only 4% coverage at >80% identity
```

---

**[Back to index.](../README.md)**
