---
title: 'Chapter 3: Simon Growth Curve'
author: "VTO"
date: "2025-03-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Data analysis for manuscript "The pQBR mercury resistance plasmids: a model set of sympatric environmental mobile genetic elements". Code written by Victoria T Orr (Victoria.Orr@liverpool.ac.uk), and James P.J. Hall and Prajwal Vishwanath Bharadwaj.

## Extracting data from multi-read platereader outputs

These first lines should be run in bash, or in the 'Terminal' window in Rstudio.

```{bash, eval=FALSE}
grep -a -A 9 "Cycles / Well" rawdata/Growth_25-10-24.csv | grep -a -v "Cycles / Well"> tmp/Growth_pQBR57.csv
```

These subsequent lines are run in R.

You can match up the wells with information about their contents. Make another table called `ARRANGEMENT.csv` (or whatever you like, but modify the following code accordingly). Format `ARRANGEMENT.csv` so that one column is called `well` and contains one row for each well A1-H12. Other columns can be titled and include whatever information you like.

Code edited

```{r }

library(tidyverse)
library(ggplot2)
library(gcplyr)
library(patchwork)

# -----------------------------------------------------------------
# Input File Names
exptdat <- read.csv("tmp/Growth_pQBR57.csv", header=FALSE, sep=",")
arrangement <- read.csv("tmp/Arrangement_pQBR57.csv", header=TRUE, sep=",")
parent_in_this_file <- seq(1, 28, by=4)
# -----------------------------------------------------------------

exptdat_list <- list()

for(j in 1:96){
  i = ((j-1)*10)+1
  well = exptdat[i,"V1"]
  welldata = data.frame(t(exptdat[(i+1):(i+8),c(2:140)]))
  colnames(welldata) <-  c("time","temp","mean","stdev","m01","m11","m10","m00")
  welldata$cycle <- c(1:nrow(welldata))
  welldata$well <- well
  exptdat_list[[j]] <- welldata
}

exptdat_long <- bind_rows(exptdat_list) %>%
  rowwise() %>%
  mutate(median_value = median(c(m01, m11, m10, m00), na.rm=TRUE))


exptdat_long$row <- substr(exptdat_long$well,1,1)
exptdat_long$col <- as.numeric(gsub("[A-H]","",exptdat_long$well))

exptdat_long <- exptdat_long %>% left_join(arrangement, by="well")

sorder <- 1:200
exptdat_long.1 <- subset(exptdat_long, strain!="blank")
exptdat_long.1 <- exptdat_long.1 %>% mutate(hours = time/3600)

exptdat_summ <- exptdat_long.1 %>%
  group_by(hours,plasmid) %>%
  summarise(OD = mean(median_value))

exptdat_long.1 <- exptdat_long.1 %>% mutate(OD = median_value)

select_medians<- filter(exptdat_long, strain == "blank" & cycle == 100 & median_value > 0.5) #taking mean measurement for non-growing blanks over the first hour as a measure of the blank.

blank_measure <- filter(exptdat_long, strain == "blank" & col != 4 & cycle < 5) %>% pull(median_value) %>% mean()

exptdat_long.1 <- mutate(exptdat_long.1, 
                         od600_corr = median_value - blank_measure) %>%
  select(hours, well, row, col, strain, parent, plasmid, replicate, od600_corr)


```

 Pre-processing using `gcplyr`
 De-noising the data

## Data has been smoothed using:
##    moving-average stored in `ods1` 
##    moving-average and moving-media stored in `ods2`
```{r}
dat <- mutate(group_by(exptdat_long.1, well, strain, plasmid),
              ods1 = smooth_data(x=hours, y = od600_corr, sm_method = "moving-median",
                                     window_width_n = 5),
              ods2 = smooth_data(x = hours, y = od600_corr, sm_method = "moving-average",
                                 window_width_n = 5))

dat <- dat %>% mutate(ods = ods2)
```

Calculating Growth Rate

This calculates the slope of the original density data and the per-capita growth rate.

```{r}

dat1 <-dat %>% mutate(ods_adj = ifelse(ods <= 0, 1e-30, ods))

dat1 <-dat1 %>% group_by(well, strain, plasmid) %>% 
  mutate(sidervi = calc_deriv(x=hours, y=ods_adj, blank=0, window_width_n = 5),
         percap3 = calc_deriv(x=hours, y=ods_adj, percapita=TRUE, blank=0,
                              window_width_n = 3, trans_y='log'),
         doub_time = doubling_time(y = percap3))



dat1 %>% filter(ods_adj > 0.01) %>% ggplot(aes(x=hours, y=percap3)) +
   geom_line() +
    ylim(-0.2, 0.2) +
   facet_wrap(~well)
```

Plotting Graphs for the four Growth Metrics

```{r}
dat_gg <- dat1 %>% filter(ods > 0.01) %>%
  group_by(well, strain, plasmid, replicate) %>%
  summarize(lag_time = lag_time(y = ods_adj, x = hours, deriv = percap3),
            max_percap = max_gc(percap3),
            max_percap_time = hours[which_max_gc(percap3)],
            max_percap_dens = ods_adj[which_max_gc((percap3))],
            min_dens = first_minima(ods_adj, return="y"),
            auc = auc(x=hours, y=ods_adj),
            max_dens = max_gc(ods_adj, na.rm = TRUE),
            max_time = extr_val(hours, which_max_gc(ods_adj)),
            thresh = first_above(y=ods_adj, x=hours,
                                 threshold = 0.75, return = "x")
  )

  dat_gg$strain <- as.factor(dat_gg$strain)
  dat_gg <- as.data.frame(dat_gg)
  
dat_gg <- dat_gg %>% filter(strain != "Free", strain != "pQBR57_L")

  ## Max Growth Rate
 ggplot(data=dat_gg, aes(x=strain, y=max_percap , fill=strain)) +
      geom_boxplot() +
    geom_point(position=position_jitterdodge(dodge.width = 0), 
               pch=21, aes(fill=replicate), show.legend = F) +
    theme_bw()+
    theme(legend.position="none", plot.title = element_text(hjust = 0.48,size=45),
          axis.title=element_text(size=30),axis.text=element_text(size=25))+
    ggtitle('Maximum Per Capita Growth Rate')+
    labs(y="Growth Rate", x="Plasmid")
  
   ggsave("MaxGrowth.png",dpi=500,width=15,height=10)
   
  ## Lagtime
 ggplot(data=dat_gg, aes(x=strain, y=lag_time, fill=strain)) +
      geom_boxplot() +
    geom_point(position=position_jitterdodge(dodge.width = 0), 
               pch=21, aes(fill=replicate), show.legend = F) +
    ggtitle('Lag Time') +
    labs(y="Lag Time", x="Plasmid")+
    theme_bw()+
    theme(legend.position="none", plot.title = element_text(hjust = 0.48,size=45),
          axis.title=element_text(size=30),axis.text=element_text(size=25))
  
 ggsave("LagTime.png",dpi=500,width=15,height=10)
  
  ## AUC
  dat_gg2 <- dat_gg
 dat_gg2 <- dat_gg2 %>%
  mutate(Plas_group = recode(strain,
                            "Free_Tn" = 1,
                            "pQBR103" = 2,
                            "pQBR102" = 3,
                            "pQBR150" = 3,
                            "pQBR30" = 3,
                            "pQBR56" = 3,
                            "pQBR57" = 3))
 
 ggplot(data=dat_gg2, aes(x=fct_reorder(strain,Plas_group), y=auc, fill=strain)) +
      geom_boxplot(outliers = FALSE) +
    geom_point(position=position_jitterdodge(dodge.width = 0), 
               pch=21, aes(fill=replicate), show.legend = F) +
    theme_bw()+
    labs(y="Area Under the Curve", x="Plasmid")+
    ggtitle('Area Under The Curve') +
    theme(legend.position="none", plot.title = element_text(hjust = 0.48,size=45),
          axis.title=element_text(size=30),axis.text=element_text(size=25))
  
 ggsave("AUC_nooutlier.png",dpi=500,width=15,height=10)
 
## Max Growth Density
  ggplot(data=dat_gg, aes(x=strain, y=max_dens, fill=strain)) +
      geom_boxplot(outliers = FALSE) +
    geom_point(position=position_jitterdodge(dodge.width = 0), 
               pch=21, aes(fill=replicate), show.legend = F) +
    ggtitle('Maximum Growth Density') +
    theme_bw()+
    theme(legend.position="none", plot.title = element_text(hjust = 0.48,size=45),
          axis.title=element_text(size=30),axis.text=element_text(size=25))+
    labs(y="Growth Density",x="Plasmid")
  
   ggsave("MaxGrowthDens_nooutlier.png",dpi=500,width=15,height=10)

```

```{r}
col3 <- "#FE6100"
col5<- "#785EF0"
col7<-"#18B6DA"
COLS<-c(col3,col5,col7)
exptdat_clean <- dat_1 %>%
  group_by(hours, strain, plasmid) %>%
  summarise( mean_OD = mean(od600_corr),
             std = sd(od600_corr)
             )   

curve_102 <- exptdat_clean %>% filter(strain %in% c("Free_Tn", "pQBR102"))
curve_103 <- exptdat_clean %>% filter(strain %in% c("Free_Tn", "pQBR103"))
curve_150 <- exptdat_clean %>% filter(strain %in% c("Free_Tn", "pQBR150"))
curve_30 <- exptdat_clean %>% filter(strain %in% c("Free_Tn", "pQBR30"))
curve_56 <- exptdat_clean %>% filter(strain %in% c("Free_Tn", "pQBR56"))
curve_57 <- exptdat_clean %>% filter(strain %in% c("Free_Tn", "pQBR57"))
curve_57_L <- exptdat_clean %>% filter(strain %in% c("Free_Tn", "pQBR57_L"))

A <- ggplot(data=curve_102, 
        aes(x=hours, y=mean_OD, color=plasmid, group=strain, fill=plasmid)) +
   geom_line(size=1) +
   geom_ribbon(aes(ymin=mean_OD-std, ymax=mean_OD+std), alpha=0.2, color=NA) +
   scale_fill_manual(values =COLS,labels=c("Free","Plasmid"))+
  scale_colour_manual(values=COLS,labels=c("Free","Plasmid"))+
  theme_bw()+
   theme(axis.line = element_line(color = "black"),
         axis.text=element_text(size=20),
        axis.title=element_text(size=25),
       # axis.title=element_blank(),
        legend.text=element_text(size=25),
        legend.title=element_blank(),
        axis.text.x=element_text(angle=45,hjust = 1)) +
 #  ymax(1.5) +
   labs(x="Time (in hours)", y="Optical Density at 600nm")+
  ggtitle("pQBR102")
#ggsave("Growth_pQBR102.png",dpi=500,width=10,height=10)

B <- ggplot(data=curve_103, 
        aes(x=hours, y=mean_OD, color=plasmid, group=strain, fill=plasmid)) +
   geom_line(size=1) +
   geom_ribbon(aes(ymin=mean_OD-std, ymax=mean_OD+std), alpha=0.2, color=NA) +
    # scale_x_continuous(breaks = mybreaks, labels = mylabs) +
   scale_fill_manual(values =COLS,labels=c("Free","Plasmid"))+
  scale_colour_manual(values=COLS,labels=c("Free","Plasmid"))+
   theme_bw()+
   theme(axis.line = element_line(color = "black"),
         axis.text=element_text(size=20),
        axis.title=element_text(size=25),
      #  axis.title=element_blank(),
        legend.text=element_text(size=25),
        legend.title=element_blank(),
        axis.text.x=element_text(angle=45,hjust = 1))+
#   ylim(0, 1.5) +
   labs(x="Time (in hours)", y="Optical Density at 600nm")+
  ggtitle("pQBR103")
#ggsave("Growth_pQBR103.png",dpi=500,width=10,height=10)

C <- ggplot(data=curve_150, 
        aes(x=hours, y=mean_OD, color=plasmid, group=strain, fill=plasmid)) +
   geom_line(size=1) +
   geom_ribbon(aes(ymin=mean_OD-std, ymax=mean_OD+std), alpha=0.2, color=NA) +
  #   scale_x_continuous(breaks = mybreaks, labels = mylabs) +
   scale_fill_manual(values =COLS,labels=c("Free","Plasmid"))+
  scale_colour_manual(values=COLS,labels=c("Free","Plasmid"))+
    theme_bw()+
   theme(axis.line = element_line(color = "black"),
         axis.text=element_text(size=20),
        axis.title=element_text(size=25),
       # axis.title=element_blank(),
        legend.text=element_text(size=25),
        legend.title=element_blank(),
        axis.text.x=element_text(angle=45,hjust = 1))+
#   ylim(0, 1.5) +
   labs(x="Time (in hours)", y="Optical Density at 600nm")+
  ggtitle("pQBR150")
#ggsave("Growth_pQBR150.png",dpi=500,width=10,height=10)

D <- ggplot(data=curve_30, 
        aes(x=hours, y=mean_OD, color=plasmid, group=strain, fill=plasmid)) +
   geom_line(size=1) +
   geom_ribbon(aes(ymin=mean_OD-std, ymax=mean_OD+std), alpha=0.2, color=NA) +
 #    scale_x_continuous(breaks = mybreaks, labels = mylabs) +
   scale_fill_manual(values =COLS,labels=c("Free","Plasmid"))+
  scale_colour_manual(values=COLS,labels=c("Free","Plasmid"))+
    theme_bw()+
   theme(axis.line = element_line(color = "black"),
         axis.text=element_text(size=20),
        axis.title=element_text(size=25),
        #axis.title=element_blank(),
        legend.text=element_text(size=25),
        legend.title=element_blank(),
        axis.text.x=element_text(angle=45,hjust = 1))+
#   ylim(0, 1.5) +
   labs(x="Time (in hours)", y="Optical Density at 600nm")+
  ggtitle("pQBR30")
#ggsave("Growth_pQBR30.png",dpi=500,width=10,height=10)

E <- ggplot(data=curve_56, 
        aes(x=hours, y=mean_OD, color=plasmid, group=strain, fill=plasmid)) +
   geom_line(size=1) +
   geom_ribbon(aes(ymin=mean_OD-std, ymax=mean_OD+std), alpha=0.2, color=NA) +
 #    scale_x_continuous(breaks = mybreaks, labels = mylabs) +
   scale_fill_manual(values =COLS,labels=c("Free","Plasmid"))+
  scale_colour_manual(values=COLS,labels=c("Free","Plasmid"))+
   theme_bw()+
   theme(axis.line = element_line(color = "black"),
         axis.text=element_text(size=20),
        axis.title=element_text(size=25),
      #  axis.title=element_blank(),
        legend.text=element_text(size=25),
        legend.title=element_blank(),
        axis.text.x=element_text(angle=45,hjust = 1))+
  # ylim(0, 1.5) +
   labs(x="Time (in hours)", y="Optical Density at 600nm")+
  ggtitle("pQBR56")
#ggsave("Growth_pQBR56.png",dpi=500,width=10,height=10)

F <- ggplot(data=curve_57, 
        aes(x=hours, y=mean_OD, color=plasmid, group=strain, fill=plasmid)) +
   geom_line(size=1) +
   geom_ribbon(aes(ymin=mean_OD-std, ymax=mean_OD+std), alpha=0.2, color=NA) +
#     scale_x_continuous(breaks = mybreaks, labels = mylabs) +
   scale_fill_manual(values =COLS,labels=c("Free","Plasmid"))+
  scale_colour_manual(values=COLS,labels=c("Free","Plasmid"))+
  theme_bw()+
  theme(axis.line = element_line(color = "black"),
         axis.text=element_text(size=20),
        axis.title=element_text(size=25),
      #  axis.title=element_blank(),
        legend.text=element_text(size=25),
        legend.title=element_blank(),
        axis.text.x=element_text(angle=45,hjust = 1))+
#   ylim(0, 1.5) +
   labs(x="Time (in hours)", y="Optical Density at 600nm")+
  ggtitle("pQBR57")
#ggsave("Growth_pQBR57.png",dpi=500,width=10,height=10)


patch1 <- (F+D+E+A+C+B) +plot_layout(ncol = 3, nrow = 2, axes = "collect", axis_titles = "collect",guides="collect")
ggsave("Growth_all.png",dpi=500,width=15,height=10)


```

Calculating statistics from growth curves.

```{r}
library(multcomp)
library(FSA)

# Maximum Growth Rate

# Running a linear model on dat_gg
dat_gg$strain <- as.factor(dat_gg$strain)
mod1 <- lm(max_percap  ~ strain, data=dat_gg)

aovres1 <- aov(data=dat_gg, max_percap  ~ strain)
tukey1 <- cld(glht(aovres1, linfct = mcp(strain = "Tukey")), decreasing = TRUE)
tukey1$mcletters$Letters

#lag time
mod2 <- lm(lag_time ~ strain, data=dat_gg)
aovres2 <- aov(data=dat_gg, lag_time ~ strain)
tukey2 <- cld(glht(aovres2, linfct = mcp(strain = "Tukey")), decreasing = TRUE)
tukey2$mcletters$Letters

# AUC
mod3 <- lm(auc ~ strain, data=dat_gg)
aovres3 <- aov(data=dat_gg, auc ~ strain)
tukey3 <- cld(glht(aovres3, linfct = mcp(strain = "Tukey")), decreasing = TRUE)
tukey3$mcletters$Letters

#Growth Density
kruskal.test(data=dat_gg, max_dens ~ strain)
dunnTest(dat_gg$max_dens, dat_gg$strain, method = "bonferroni")
