---
title: 'pQBR PLasmids
author: "VTO"
date: "2025-03-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Data analysis for manuscript "The pQBR mercury resistance plasmids: a model set of sympatric environmental mobile genetic elements". Code written by Victoria T Orr (Victoria.Orr@liverpool.ac.uk), supervised by James P.J. Hall. 

Prerequisite file:

-"Hybrid_data.csv" - data collected from conjguation/fitness experiments (excluding growth curves)

Packages and colour scheme:
```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(nlme)
library(forcats)
library(ggtext)
library(patchwork)

pd <- position_dodge(width=0.5)
col1 <- "#FFB000"
col2 <- "#DC267F"
col4 <- "#FE6100"
col3<- "#785EF0"
col5<- "#000000"
col6<-"#0101DF"
col7<-"#18B6DA"
COL=c(col4,col7)

```

Hybrid_data.csv consists of several datasets on differnt dates. As Nov_23 yields unusual counts compared with other datasets,therefore we chose to exclude. 

We chose to include Dec24 and Jan25 datasets due to small difference in time conducted, with the exception of pQBR55 (Oct-22) as this was the only dataset for this plasmid and preliminary testing had yielded a consistent conjugation rate and lack of transposon mobilisation detected in this assay. pQBR56 and pQBR51 were excluded due to minimal datapoints and genetic similarity to pQBR57 and pQBR150.

```{r}
conj_exp<-read.csv("Hybrid_data.csv",header=TRUE,sep=",")

`%notin%` <- Negate(`%in%`)

conj_exp<- conj_exp %>% filter(Plas_group != "Con") %>% 
  mutate(Start_D=(Start_DC*10^(Dilution_S))/Quantity_S,
         Start_R=(Start_RC*10^(Dilution_S))/Quantity_S,
         Start_N=(Start_NC*10^(Dilution_S))/Quantity_S,
         End_D=(End_DC*10^(Dilution_E))/Quantity_E,
         End_R=(End_RC*10^(Dilution_E))/Quantity_E,
         End_N=(End_NC*10^(Dilution_E))/Quantity_E,
         End_D=(End_DC*10^(Dilution_E))/Quantity_E,
         Plas_T=(Plas_TC*10^(Dilution_P))/Quantity_P,
         Tran_T=(Tran_TC*10^(Dilution_T))/Quantity_T,
         psi = log(End_N/Start_N)/24,
         gamma.p = psi * log(1+((Plas_T/End_R)*(End_N/End_D))) * (1/(End_N-Start_N)),
         gamma.t = psi * log(1+((Tran_T/End_R)*(End_N/End_D))) * (1/(End_N-Start_N)))

conj_exp_filtered<- conj_exp %>% 
  select(Label,Plasmid,Plas,Plas_group,Strain_Rep,Plas_Rep, gamma.p, gamma.t, Dat_rep, Data) %>%
  pivot_longer(cols=c(gamma.p, gamma.t), names_to = "type", values_to = "gamma") %>%
  filter(Data %notin% "Nov_23") %>%
  filter(Plasmid %notin% c("pQBR56", "pQBR57", "pQBR51")) %>%
  filter(!(Data == "Oct_22" & Plasmid != "pQBR55_L")) %>%
  mutate(Plasmid = case_when(
    Plasmid == "pQBR57_L" ~ "pQBR57",
    Plasmid == "pQBR55_L" ~ "pQBR55",  
    Plasmid == "pQBR103_L" ~ "pQBR103",
    Plasmid == "pP19E3_L" ~ "pP19E3",
    TRUE ~ Plasmid 
  )) 

Limits <- filter(conj_exp_filtered, Plasmid == "LIM") %>%
  filter(type=="gamma.t")

conj_exp_filtered <-conj_exp_filtered %>%
  filter(Plas_Rep != "13") %>%
  group_by(Plasmid, Plas_Rep) %>%
  filter(!any(is.na(gamma))) %>%
  ungroup()

conj_exp_filtered <- conj_exp_filtered %>%
  group_by(Plasmid) %>%
  mutate(plas_label = dense_rank(Plas)) %>%
  ungroup()

conj_summ <- conj_exp_filtered %>%
     group_by(Plasmid,Plas_group,type) %>%
    summarise(gamma = mean(gamma, na.rm=TRUE))

pd <- position_dodge(width=0.5)

conj_exp_log <- conj_exp_filtered %>%
  mutate(conj_exp_filtered, log_gamma = log10(gamma))

conj_exp_log <- conj_exp_log %>%
   mutate(log_gamma = replace(log_gamma, log_gamma == -Inf, -21))

conj_summstat <- conj_exp_log %>%
  group_by(Plasmid, type) %>%
  summarise(
    mean = mean(log_gamma, na.rm=TRUE),
    std = sd(log_gamma, na.rm = TRUE)) %>%
  left_join(conj_exp_log %>% select(Plasmid,
                              type,Dat_rep,Plas_group) %>%
              distinct(), by = c("Plasmid", "type")) %>%
  mutate(ymin = mean - std, ymax = mean + std)

conj_exp_log$Plas_group <- as.numeric(conj_exp_log$Plas_group)

conj_summstat<- rename(conj_summstat, "log_gamma"="mean")

plasgroup_colors <- c(
  "1" = "#f68600ff", #GroupII
  "2" = "#5900abff", #GroupIII
  "3" = "#00ab2fff", #GroupI
  "4" = "#00ab2fff", #GroupI
  "5" = "#AA8800", #pP19E3
  "6" = "#00a9abff" #GroupIV
  # Add all Plas_group levels and their desired colors
)
label_lookup <- unique(conj_exp_log[, c("Plasmid", "Plas_group")])

label_lookup$Plasmid_col <- paste0(
  "<span style='color:", 
  plasgroup_colors[label_lookup$Plas_group], 
  "'>", 
  label_lookup$Plasmid, 
  "</span>")

conj_exp_log$Plasmid_col <- NULL
conj_exp_log <- merge(conj_exp_log, label_lookup, by = c("Plasmid", "Plas_group"))
conj_summstat$Plasmid_col <- NULL
conj_summstat <- merge(conj_summstat, label_lookup, by = "Plasmid")

conj_exp_log$Plasmid_col <- 
  fct_reorder(conj_exp_log$Plasmid_col, conj_exp_log$Plas_group)
conj_summstat$Plasmid_col <-
  factor(conj_summstat$Plasmid_col,
         levels = levels(conj_exp_log$Plasmid_col))

pd2=position_jitter(height=0, width=0.15)

ggplot(conj_exp_log, aes(x=Plasmid_col, y=log_gamma,
                   ymax=log_gamma,
                   colour=type, fill=type,
                   shape=as.factor(plas_label))) +
  geom_point(position=pd2, size=4) +
  theme_bw() +
  theme(
    axis.text=element_text(size=20),
    axis.title=element_text(size=25),
    legend.text=element_text(size=25),
    legend.title=element_blank(),
    axis.text.x=element_markdown(angle=45, hjust=1),  
    legend.position = "bottom",
    panel.grid = element_blank()) +
  labs(x="Plasmid", y=expression(log[10](gamma))) +
  scale_colour_manual(values = COL,
                      labels=c("Plasmid transfer", "Transposon transfer")) +
  scale_fill_manual(values = COL,
                    labels=c("Plasmid transfer", "Transposon transfer")) +
  scale_shape_manual(values = c(0,16,2,3,18,4,7,8,9,11)) +
  guides(shape = "none") +
  geom_pointrange(data=conj_summstat,
                  aes(x=Plasmid_col, ymin=ymin, ymax=ymax, shape=NULL),
                  size=2, alpha=0.3) + 
  scale_y_continuous(limits = c(min(conj_exp_log$log_gamma-1,
                                    na.rm = TRUE), NA), expand = c(0, 0)) +
  geom_hline(yintercept=log10(Limits$gamma), linetype='dotted', col='black')

ggsave("Hybrid_paper.png",dpi=1200,width=15,height=10)


```

Analysing conjugation/transposon transfer rates - updated with refined dataset for preprint
```{r}

lod <-Limits$gamma

conj_m <- conj_exp_log %>% select(Label,Plasmid,Plas,plas_label,Plas_group,
                                  Strain_Rep,Plas_Rep,Dat_rep,type,gamma) %>%
  pivot_wider(names_from = "type", values_from = "gamma") %>%
  mutate(gamma.t_adj = ifelse(gamma.t < lod, lod, gamma.t),
         ratio = (gamma.t_adj/gamma.p),
         ratio_log = log10(ratio))

conj_m$ratio_group <- ifelse(conj_m$gamma.t_adj < 6e-21, 
                             "Positive", 
                             "Negative")
conj_m_summ <- conj_m %>%
  group_by(Plasmid) %>%
  summarise(
    log_ratio = mean(ratio_log),
    std = sd(ratio_log)) %>%
  left_join(label_lookup %>% select(Plasmid, Plasmid_col) %>%
  distinct(), by = c("Plasmid")) %>%
  mutate(ymin = log_ratio - std, ymax = log_ratio + std)

conj_m$Plasmid_col <- NULL
conj_m <- merge(conj_m, label_lookup, by = c("Plasmid", "Plas_group"))
conj_m_summ$Plasmid_col <- NULL
conj_m_summ <- merge(conj_m_summ, label_lookup, by = "Plasmid")

conj_m$Plas_group <- as.numeric(as.character(conj_m$Plas_group))

conj_m <-conj_m %>% mutate(log_ratio=ratio_log)

conj_m$Plasmid_col <- fct_reorder(conj_m$Plasmid_col,
                                  conj_m$Plas_group)
conj_m_summ$Plasmid_col <- factor(conj_m_summ$Plasmid_col,
                                  levels = levels(conj_m$Plasmid_col))

ggplot(conj_m, aes(x=Plasmid_col, y=log_ratio,
                   ymax=log_ratio, shape=as.factor(plas_label),color=ratio_group)) +
  geom_point(position=pd2, size=4) +
  theme_bw() +
  theme(
    axis.text=element_text(size=20),
    axis.title=element_text(size=25),
    legend.text=element_text(size=25),
    legend.title=element_blank(),
    axis.text.x=element_markdown(angle=45, hjust=1),  # Enable markdown for colored labels
    panel.grid = element_blank()
  ) +
  labs(x="Plasmid", y=expression(log[10](Ratio))) +
  scale_shape_manual(values = c(0,16,2,3,18,4,7,8,9,11)) +
  scale_color_manual(values = c("Positive"="purple",
                                "Negative"="black"))+
  guides(shape = "none", colour = "none") +
  geom_pointrange(data=conj_m_summ,
                  aes(x=Plasmid_col, y = log_ratio, ymin=ymin, ymax=ymax, shape=NULL),inherit.aes=FALSE,
                  size=2, alpha=0.3)

ggsave("Conj_Ratios_paper.png",dpi=1200,width=15,height=10)

conj_m$Plas_group <- factor(conj_m$Plas_group)


conj_mod1 <- lm(ratio_log ~ Plasmid, data = conj_m)
summary(conj_mod1)
TukeyHSD(aov(conj_mod1))

conj_mod2 <- lm(ratio_log ~ Plas_group, data = conj_m)
summary(conj_mod2)
TukeyHSD(aov(conj_mod2))

summary(aov(data = conj_m, gamma.t~gamma.p))

conj_mod3<- lm(gamma.p ~ Plasmid, data=conj_m)
summary(conj_mod3)
TukeyHSD(aov(conj_mod3))

conj_mod4<- lm(gamma.p ~ Plas_group, data=conj_m)
summary(conj_mod4)
TukeyHSD(aov(conj_mod4))


#Getting some exact numbers for how many tns per plasmid transferc
p57 <- conj_m %>%
  filter(Plasmid == "pQBR57") %>%
  summarise(
    plasmid_mean_log10 = mean(log10(gamma.p), na.rm = TRUE),
    plasmid_sd_log10   = sd(log10(gamma.p), na.rm = TRUE),
    transposon_mean_log10 = mean(log10(gamma.t), na.rm = TRUE),
    transposon_sd_log10   = sd(log10(gamma.t), na.rm = TRUE)
  )

p102 <- conj_m %>%
  filter(Plasmid == "pQBR102") %>%
  summarise(
    plasmid_mean_log10 = mean(log10(gamma.p), na.rm = TRUE),
    plasmid_sd_log10   = sd(log10(gamma.p), na.rm = TRUE),
    transposon_mean_log10 = mean(log10(gamma.t), na.rm = TRUE),
    transposon_sd_log10   = sd(log10(gamma.t), na.rm = TRUE)
  )

log10_ratio_57 <- p57$plasmid_mean_log10 - p57$transposon_mean_log10
sd_log10_ratio_57 <- sqrt(p57$plasmid_sd_log10^2 +
                            p57$transposon_sd_log10^2)

ratio_57 <- 10^log10_ratio_57

log10_ratio_102 <- p102$plasmid_mean_log10 - p102$transposon_mean_log10
sd_log10_ratio_102 <- sqrt(p102$plasmid_sd_log10^2 +
                             p102$transposon_sd_log10^2)

ratio_102 <- 10^log10_ratio_102

ci_57 <- 10^(log10_ratio_57 + c(-1, 1) * 1.96 * sd_log10_ratio_57)
ci_102 <- 10^(log10_ratio_102 + c(-1, 1) * 1.96 * sd_log10_ratio_102)

```

Now analysing the comparative fitness data
```{r}
fit_exp<-read.csv("Hybrid_data.csv",header=TRUE,sep=",")

fit_exp<- fit_exp %>% mutate(Start_D=(Start_DC*10^(Dilution_S))/Quantity_S,
         Start_R=(Start_RC*10^(Dilution_S))/Quantity_S,
         Start_N=(Start_NC*10^(Dilution_S))/Quantity_S,
         End_D=(End_DC*10^(Dilution_E))/Quantity_E,
         End_R=(End_RC*10^(Dilution_E))/Quantity_E,
         End_N=(End_NC*10^(Dilution_E))/Quantity_E,
         End_D=(End_DC*10^(Dilution_E))/Quantity_E,
         Plas_T=(Plas_TC*10^(Dilution_P))/Quantity_P,
         Tran_T=(Tran_TC*10^(Dilution_T))/Quantity_T,
         psi = log(End_N/Start_N)/24,
         gamma.p = psi * log(1+((Plas_T/End_R)*(End_N/End_D))) * (1/(End_N-Start_N)),
         gamma.t = psi * log(1+((Tran_T/End_R)*(End_N/End_D))) * (1/(End_N-Start_N)))

comp1 <- fit_exp %>%
  mutate(Mlac = (log(End_R/(Start_R/100))/24)) %>%
  mutate(Mvs = (log(End_D/(Start_D/100))/24))

#calculating fitness - w
comp2 <- comp1 %>%
  mutate(w = Mvs/Mlac)
comp2.5 <- comp2 %>%
  group_by(Plasmid) %>%
  summarise(mean=mean(w),std=sd(w))%>%
  mutate(ymin = mean - std,
         ymax = mean + std)

con_mean <-   1.018404 

con <- subset(comp2, Plasmid =="Con")

t.test(con$w, mu=1, data=con)

comp3<- comp2 %>% mutate(w_corr = w/con_mean)
comp3[comp3 ==-Inf]<-0

comp3 <- comp3 %>% filter(Plasmid != "pQBR57_L") %>% 
  mutate(Plasmid = case_when(
    Plasmid == "pQBR55_L" ~ "pQBR55",  
    Plasmid == "pQBR103_L" ~ "pQBR103",
    Plasmid == "pP19E3_L" ~ "pP19E3",
    TRUE ~ Plasmid)) %>%
  filter(Plasmid !="Con", Plasmid !="LIM")

comp3$Plas_group <- as.numeric(comp3$Plas_group)
comp3$Plasmid <- as.factor(comp3$Plasmid)
comp3stat <- comp3 %>%
  group_by(Plasmid) %>%
  summarise(mean=mean(w_corr),
            std=sd(w_corr))%>%
  mutate(ymin = mean - std,
         ymax = mean + std)

comp3stat<- rename(comp3stat, "w_corr"="mean")

plasgroup_colors2 <- c(
  "1" = "#f68600ff", #GroupII
  "2" = "#5900abff", #GroupIII
  "3" = "#00ab2fff", #GroupI
  "4" = "#00ab2fff", #GroupI
  "5" = "#AA8800", #pP19E3
  "6" = "#00a9abff", #GroupIV
  "Con" = "black"
  # Add all Plas_group levels and their desired colors
)

label_lookup2 <- unique(comp3[, c("Plasmid", "Plas_group")])
label_lookup2$Plasmid_col <- paste0(
  "<span style='color:", 
  plasgroup_colors2[label_lookup2$Plas_group], 
  "'>", 
  label_lookup2$Plasmid, 
  "</span>")

comp3$Plasmid_col <- NULL
comp3 <- merge(comp3, label_lookup2,
               by = c("Plasmid", "Plas_group"))
comp3stat$Plasmid_col <- NULL
comp3stat <- merge(comp3stat,
                   label_lookup2, by = "Plasmid")

comp3$Plasmid_col <- fct_reorder(comp3$Plasmid_col,
                                 comp3$Plas_group)
comp3stat$Plasmid_col <- factor(comp3stat$Plasmid_col,
                                levels = levels(comp3$Plasmid_col))


pd <- position_jitterdodge(dodge.width=0.6, jitter.width = 0.1)
pd2<-position_dodge(width=0.6)

ggplot(comp3, aes(x=Plasmid_col,y=w_corr))+ 
  geom_point(position=pd2,size=4)+
  geom_hline(yintercept = 1, linetype="dashed")+
  theme_bw()+
  theme(axis.text=element_text(size=20),axis.title=element_text(size=25), legend.text=element_text(size=25),legend.title=element_text(size=25),
        axis.text.x = ggtext::element_markdown(angle = 45, hjust = 1))+labs(x="Plasmid", y="Relative Fitness (W)")+ 
  geom_pointrange(data=comp3stat,aes(x=Plasmid_col,ymin=w_corr-std,ymax=w_corr+std),size=2,alpha=0.3,position=pd2) +
   scale_y_continuous(breaks = seq(0.5, max(comp3$w_corr) + 0.2, by = 0.2))
  
ggsave("Comp_fit.png",dpi=1200,width=12,height=10)

```

Analysing fitness values
```{r}
fit <- comp3

F3 <- aov(w_corr ~ Plasmid, data = fit) 

F4<- aov(w_corr~as.factor(Plas_group), data=fit)

```
