---
title: "pQBR Plasmids: Experimental Analysis"
output: rmarkdown::github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list=ls())

library(tidyverse)
library(ggplot2)
library(dplyr)
library(nlme)
library(forcats)
library(ggtext)
library(patchwork)
library(knitr)
```

Data analysis for manuscript "The pQBR mercury resistance plasmids: a model set of sympatric environmental mobile genetic elements". Code written by Victoria T Orr ([Victoria.Orr@liverpool.ac.uk](mailto:Victoria.Orr@liverpool.ac.uk)), supervised by [James P.J. Hall](mailto:j.p.j.hall@liverpool.ac.uk). 

Prerequisite file:

- "Hybrid_data.csv" - data collected from conjguation/fitness experiments (excluding growth curves)

Functions and colour scheme:

```{r}
`%notin%` <- Negate(`%in%`)

pd <- position_dodge(width=0.5)
col1 <- "#FFB000"
col2 <- "#DC267F"
col4 <- "#FE6100"
col3 <- "#785EF0"
col5 <- "#000000"
col6 <- "#0101DF"
col7 <- "#18B6DA"
```

Hybrid_data.csv consists of several datasets collected on different dates.

For analysis, we aggregated the data from the Dec_24 and Jan_25 experiments, since these were performed just one month apart. We additionally included the pQBR55 data from Oct_22 as this was the only dataset for this plasmid, and results for the other plasmids tested (pP19E3, pQBR57, pQBR103) were consistent with those from later experiments. 

```{r}
exp_df <- read.csv("Hybrid_data.csv", header = TRUE, sep = ",", 
                     colClasses = c(rep("factor",6), rep("numeric",14),"factor","factor")) %>%
  mutate(Start_D = 0.01 * (Start_DC*10^(Dilution_S))/Quantity_S,
         Start_R = 0.01 * (Start_RC*10^(Dilution_S))/Quantity_S,
         Start_N = Start_D + Start_R,
         End_D = (End_DC*10^(Dilution_E))/Quantity_E,
         End_R = (End_RC*10^(Dilution_E))/Quantity_E,
         End_N = End_D + End_R)

conj_exp <- exp_df %>%
  filter(!(is.na(Plas_TC))) %>% # remove data without conjugation measurements
  mutate(Plas_T = (Plas_TC*10^(Dilution_P))/Quantity_P,
         Tran_T = (Tran_TC*10^(Dilution_T))/Quantity_T,
         psi = log(End_N/Start_N)/24,
         gamma.p = psi * log(1+((Plas_T/End_R)*(End_N/End_D))) * (1/(End_N-Start_N)),
         gamma.t = psi * log(1+((Tran_T/End_R)*(End_N/End_D))) * (1/(End_N-Start_N)))

# calculate minimal limits

lims <- data.frame(
  Start_N = 0.01 * (125*10^6)/0.05,
  End_N = (100*10^7)/0.05,
  End_D = (36*10^7)/0.05,
  End_R = (64*10^7)/0.05,
  Plas_T = 1*10^2/0.05, # using the least diluted plate, and most plated
  Tran_T = 1*10^0/0.2 # using the least diluted plate, and most plated
) %>% 
  mutate(psi = log(End_N/Start_N)/24,
         gamma.p = psi * log(1+((Plas_T/End_R)*(End_N/End_D))) * (1/(End_N-Start_N)),
         gamma.t = psi * log(1+((Tran_T/End_R)*(End_N/End_D))) * (1/(End_N-Start_N)))

# plot

ggplot(data=conj_exp, aes(x=Plasmid, shape=Plas)) + 
  geom_point(aes(y=log10(gamma.p)), colour="darkred") + 
  geom_point(aes(y=log10(gamma.t)), colour="darkblue") + 
  facet_grid(.~Data) + theme(axis.text.x=element_text(angle=45, hjust=1))

ggplot(data=conj_exp, aes(x=Plasmid, colour=Data)) + 
  geom_point(aes(y=log10(gamma.p)), shape=16, position=pd) + 
  geom_point(aes(y=log10(gamma.t)), shape=17, position=pd) + 
  theme(axis.text.x=element_text(angle=45, hjust=1))
```

Filter and wrangle the dataset for analysis. 

```{r}
conj_exp_filtered <- conj_exp %>% 
  select(Label, Plasmid, Plas, Plas_group, Strain_Rep, 
         Plas_Rep, gamma.p, gamma.t, Dat_rep, Data) %>%
  pivot_longer(cols=c(gamma.p, gamma.t), names_to = "type", values_to = "gamma") %>%
  filter(!(Data == "Oct_22" & Plasmid != "pQBR55_L")) %>%
  mutate(Plasmid = case_when(
    Plasmid == "pQBR57_L" ~ "pQBR57",
    Plasmid == "pQBR55_L" ~ "pQBR55",  
    Plasmid == "pQBR103_L" ~ "pQBR103",
    Plasmid == "pP19E3_L" ~ "pP19E3",
    TRUE ~ Plasmid 
  )) 

Limits <- pivot_longer(lims, cols=c(gamma.p, gamma.t), names_to = "type", values_to = "gamma") %>%
  filter(type=="gamma.t")

conj_exp_filtered <- conj_exp_filtered %>%
  group_by(Plasmid, Plas_Rep) %>%
  filter(!any(is.na(gamma))) %>%
  ungroup() %>%
  group_by(Plasmid) %>%
  mutate(plas_label = dense_rank(Plas)) %>%
  ungroup()

pd <- position_dodge(width=0.5)

conj_exp_log <- conj_exp_filtered %>%
  mutate(log_gamma = log10(gamma),
         log_gamma = replace(log_gamma, log_gamma == -Inf, -20))

conj_summ <- conj_exp_log %>%
  group_by(Plasmid,Plas_group,type) %>%
  summarise(mean = mean(gamma, na.rm=TRUE),
            mean_log = mean(log_gamma),
            sd = sd(log_gamma)) %>%
  rename(log_gamma = mean_log)
```

Calculate summary statistics and plot.

```{r}
plasgroup_colours <- data.frame(
  Plas_group = c("1","2","3","5","6"),
  colour = c("#f68600ff", #GroupII
             "#5900abff", #GroupIII
             "#00ab2fff", #GroupI
             "#AA8800", #pP19E3
             "#00a9abff") #GroupIV
  # Add all Plas_group levels and their desired colors
)

label_lookup <- unique(conj_exp_log[, c("Plasmid", "Plas_group")]) %>%
  left_join(plasgroup_colours, by="Plas_group") %>%
  mutate(Plasmid_col = paste0("<span style='color:", 
                              colour, 
                              "'>", 
                              Plasmid, 
                              "</span>")) %>%
  select(-colour)

# conj_exp_log$Plasmid_col <- NULL
conj_exp_log <- merge(conj_exp_log, label_lookup, by = c("Plasmid", "Plas_group")) %>%
  mutate(Plasmid_col = fct_reorder(Plasmid_col, as.numeric(Plas_group)))
# conj_summstat$Plasmid_col <- NULL
conj_summ <- merge(conj_summ, label_lookup, by = "Plasmid") %>%
  mutate(Plasmid_col = factor(Plasmid_col,
                              levels = levels(conj_exp_log$Plasmid_col)))
```

Plot.

```{r}
pd2 <- position_jitter(height=0, width=0.15)

ggplot(conj_exp_log, aes(x=Plasmid_col, y=log_gamma,
                   ymax=log_gamma,
                   colour=type, fill=type,
                   shape=as.factor(plas_label))) +
  geom_point(position=pd2, size=4) +
  theme_bw() +
  theme(
    axis.text=element_text(size=20),
    axis.title=element_text(size=25),
    legend.text=element_text(size=25),
    legend.title=element_blank(),
    axis.text.x=element_markdown(angle=45, hjust=1),  
    legend.position = "bottom",
    panel.grid = element_blank()) +
  labs(x="Plasmid", y=expression(log[10](gamma))) +
  scale_colour_manual(values = c(col4, col7),
                      labels=c("Plasmid transfer", "Transposon transfer")) +
  scale_fill_manual(values = c(col4, col7),
                    labels=c("Plasmid transfer", "Transposon transfer")) +
  scale_shape_manual(values = c(0,16,2,3,18,4,7,8,9,11)) +
  guides(shape = "none") +
  geom_pointrange(data=conj_summ,
                  aes(x=Plasmid_col, ymin=log_gamma-sd, ymax=log_gamma+sd, shape=NULL),
                  size=2, alpha=0.3) + 
  scale_y_continuous(limits = c(min(conj_exp_log$log_gamma-1,
                                    na.rm = TRUE), NA), expand = c(0, 0)) +
  geom_hline(yintercept=log10(Limits$gamma), linetype='dotted', col='black')

ggsave("Hybrid_paper.png",dpi=1200,width=15,height=10)
```

Analysing conjugation/transposon transfer rates - updated with refined dataset for preprint

```{r}
lod <-Limits$gamma

conj_m <- conj_exp_log %>% select(Label,Plasmid,Plas,plas_label,Plas_group,Plasmid_col,
                                  Strain_Rep,Plas_Rep,Dat_rep,Data,type,gamma) %>%
  pivot_wider(names_from = "type", values_from = "gamma") %>%
  mutate(gamma.t_adj = ifelse(gamma.t < lod, lod, gamma.t),
         ratio = (gamma.t_adj/gamma.p),
         log_ratio = log10(ratio))

conj_m$ratio_group <- ifelse(conj_m$gamma.t < 6e-21, 
                             "Positive", 
                             "Negative")

conj_m_summ <- conj_m %>%
  group_by(Plasmid, Plas_group) %>%
  summarise(mean = mean(log_ratio),
            std = sd(log_ratio)) %>%
  rename(log_ratio = mean) %>%
  left_join(label_lookup %>% select(Plasmid, Plasmid_col) %>%
              distinct(), by = c("Plasmid")) %>%
  mutate(ymin = log_ratio - std, ymax = log_ratio + std)

ggplot(conj_m, aes(x=Plasmid_col, y=log_ratio,
                   ymax=log_ratio, shape=as.factor(plas_label),color=ratio_group)) +
  geom_point(position=pd2, size=4) +
  theme_bw() +
  theme(
    axis.text=element_text(size=20),
    axis.title=element_text(size=25),
    legend.text=element_text(size=25),
    legend.title=element_blank(),
    axis.text.x=element_markdown(angle=45, hjust=1),  # Enable markdown for colored labels
    panel.grid = element_blank()
  ) +
  labs(x="Plasmid", y=expression(log[10](Ratio))) +
  scale_shape_manual(values = c(0,16,2,3,18,4,7,8,9,11)) +
  scale_color_manual(values = c("Positive"="purple",
                                "Negative"="black"))+
  guides(shape = "none", colour = "none") +
  geom_pointrange(data=conj_m_summ,
                  aes(x=Plasmid_col, y = log_ratio, ymin=ymin, ymax=ymax, shape=NULL),inherit.aes=FALSE,
                  size=2, alpha=0.3)

ggsave("Conj_Ratios_paper.png",dpi=1200,width=15,height=10)
```

Statistical tests. First run through all the data together.

```{r}
# ratio depends on the plasmid
conj_mod1 <- lm(log_ratio ~ Plasmid, data = conj_m)
plot(conj_mod1)
summary(conj_mod1)
anova(conj_mod1) #F(8,62) = 3.34, p < 1e-15
TukeyHSD(aov(conj_mod1))

# ratio depends on the plasmid group
conj_mod2 <- lm(log_ratio ~ Plas_group, data = conj_m)
plot(conj_mod2)
summary(conj_mod2)
anova(conj_mod2) #F(4,66) = 28.3, p < 1e-12
TukeyHSD(aov(conj_mod2))

# directly compare conjugation rate and transposon transfer rates

ggplot(conj_m, aes(x=log10(gamma.p), y=log10(gamma.t_adj))) + 
  geom_point(aes(colour=Plasmid)) + geom_smooth(method="lm")

cor.test(x=conj_m$gamma.p, y=conj_m$gamma.t_adj, method="spearman")
# significant correlation, 0.463, p < 1e-4
# but these data include pQBR103 and pQBR55 that had rare/no transposon mobilisation and low conjugation rates
# remove and retest

conj_m_flt <- conj_m %>% filter(!(Plasmid %in% c("pQBR103","pQBR55")))

cor.test(x=conj_m_flt$gamma.p, 
         y=conj_m_flt$gamma.t_adj, method="spearman")

# rho = 0.19, p = 0.167
# nb can compare non-transformed values, since spearman's rho considers rank order

# conjugation rate depends on plasmid
conj_mod3<- lm(log10(gamma.p) ~ Plasmid, data=conj_m)
plot(conj_mod3)
summary(conj_mod3)
anova(conj_mod3) #F(8,62)=182, p < 1e-15
TukeyHSD(aov(conj_mod3))
```

These analyses look at the aggregated data. Are the results consistent if each experiment is analysed separately?

Main results: 

- correlation between gamma.p and gamma.t, which is lost if the low-conjugation rate plasmids are removed
- significant difference between plasmids

```{r}
ggplot(data=conj_m, aes(x=Plasmid, colour=Data)) + 
  geom_point(aes(y=log10(gamma.p)), shape=16, position=pd) + 
  geom_point(aes(y=log10(gamma.t)), shape=17, position=pd) + 
  theme(axis.text.x=element_text(angle=45, hjust=1))

conj_m_24 <- filter(conj_m_flt, Data=="Dec_24")
conj_m_25 <- filter(conj_m_flt, Data=="Jan_25")

cor.test(x=conj_m_24$gamma.p, 
         y=conj_m_24$gamma.t_adj, method="spearman")
cor.test(x=conj_m_25$gamma.p, 
         y=conj_m_25$gamma.t_adj, method="spearman")
# both of these results are nonsignificant, because pQBR55 is omitted from these experiments, as above. 

conj_mod1_24 <- lm(log_ratio ~ Plasmid, data = filter(conj_m, Data=="Dec_24"))
anova(conj_mod1_24) # significant effect of plasmid
conj_mod1_24_results <- as.data.frame(TukeyHSD(aov(conj_mod1_24))$Plasmid) %>%
  rownames_to_column("comparison") %>% mutate(star = ifelse(`p adj` < 0.05, "*", ""))

conj_mod1_25 <- lm(log_ratio ~ Plasmid, data = filter(conj_m, Data=="Jan_25"))
anova(conj_mod1_25) # significant effect of plasmid
conj_mod1_25_results <- as.data.frame(TukeyHSD(aov(conj_mod1_25))$Plasmid) %>%
  rownames_to_column("comparison") %>% mutate(star = ifelse(`p adj` < 0.05, "*", ""))

left_join(conj_mod1_24_results, conj_mod1_25_results, by="comparison",
          suffix = c(".24",".25")) %>%
  select(comparison, `p adj.24`, `p adj.25`, star.24, star.25) %>% kable()
```

Where there is a significant effect for one experiment, there's usually a significant effect for the other (and vice versa). A few exceptions, mostly because pP19E3 had only 2 replicates in the Jan_25 dataset and hence large standard error. Some minor discrepancies between the pQBR102 and pQBR57 data also (better than expected mobilisation for pQBR57 in the January dataset).

The main results thus stand when the experiments are analysed separately, justifying aggregation in the figure. 

```{r}
#Getting some exact numbers for how many tns per plasmid transfer
p57 <- conj_m %>%
  filter(Plasmid == "pQBR57") %>%
  summarise(
    plasmid_mean_log10 = mean(log10(gamma.p), na.rm = TRUE),
    plasmid_sd_log10   = sd(log10(gamma.p), na.rm = TRUE),
    transposon_mean_log10 = mean(log10(gamma.t), na.rm = TRUE),
    transposon_sd_log10   = sd(log10(gamma.t), na.rm = TRUE)
  )

p102 <- conj_m %>%
  filter(Plasmid == "pQBR102") %>%
  summarise(
    plasmid_mean_log10 = mean(log10(gamma.p), na.rm = TRUE),
    plasmid_sd_log10   = sd(log10(gamma.p), na.rm = TRUE),
    transposon_mean_log10 = mean(log10(gamma.t), na.rm = TRUE),
    transposon_sd_log10   = sd(log10(gamma.t), na.rm = TRUE)
  )

log10_ratio_57 <- p57$plasmid_mean_log10 - p57$transposon_mean_log10
sd_log10_ratio_57 <- sqrt(p57$plasmid_sd_log10^2 +
                            p57$transposon_sd_log10^2)

(ratio_57 <- 10^log10_ratio_57) #7255

log10_ratio_102 <- p102$plasmid_mean_log10 - p102$transposon_mean_log10
sd_log10_ratio_102 <- sqrt(p102$plasmid_sd_log10^2 +
                             p102$transposon_sd_log10^2)

(ratio_102 <- 10^log10_ratio_102) #2359

(ci_57 <- 10^(log10_ratio_57 + c(-1, 1) * 1.96 * sd_log10_ratio_57)) #60 to 8.74e5
(ci_102 <- 10^(log10_ratio_102 + c(-1, 1) * 1.96 * sd_log10_ratio_102)) #842 to 6604
```

Now analysing the comparative fitness data

```{r}
#calculating fitness - w
comp1 <- exp_df %>%
  mutate(Mlac = (log(End_R/(Start_R/100))/24),
         Mvs  = (log(End_D/(Start_D/100))/24),
         w = Mvs/Mlac)

con <- filter(comp1, Plasmid == "Con")

(con_mean <- con %>%
  select(w) %>% summarise(w = mean(w)) %>% pull()) # 1.009498

t.test(con$w, mu=1, data=con) # t = 0.82179, df = 9, p = 0.4324

comp3 <- comp1 %>% mutate(w_corr = w/con_mean) %>% #filter(Plasmid != "pQBR57_L") %>% 
  mutate(Plasmid = case_when(
    Plasmid == "pQBR55_L" ~ "pQBR55",  
    Plasmid == "pQBR103_L" ~ "pQBR103",
    Plasmid == "pP19E3_L" ~ "pP19E3",
    Plasmid == "pQBR57_L" ~ "pQBR57",
    TRUE ~ Plasmid)) %>%
  filter(Plasmid != "Con")

comp3$Plas_group <- as.numeric(comp3$Plas_group)
comp3$Plasmid <- as.factor(comp3$Plasmid)

comp3stat <- comp3 %>%
  group_by(Plasmid) %>%
  summarise(mean=mean(w_corr),
            std=sd(w_corr))%>%
  mutate(ymin = mean - std,
         ymax = mean + std) %>%
  rename("w_corr"="mean")

plasgroup_colors2 <- c(
  "1" = "#f68600ff", #GroupII
  "2" = "#5900abff", #GroupIII
  "3" = "#00ab2fff", #GroupI
  "4" = "#AA8800", #pP19E3
  "5" = "#00a9abff", #GroupIV
  "Con" = "black"
  # Add all Plas_group levels and their desired colors
)

label_lookup2 <- unique(comp3[, c("Plasmid", "Plas_group")])
label_lookup2$Plasmid_col <- paste0(
  "<span style='color:", 
  plasgroup_colors2[label_lookup2$Plas_group], 
  "'>", 
  label_lookup2$Plasmid, 
  "</span>")

comp3$Plasmid_col <- NULL
comp3 <- merge(comp3, label_lookup2,
               by = c("Plasmid", "Plas_group"))
comp3stat$Plasmid_col <- NULL
comp3stat <- merge(comp3stat,
                   label_lookup2, by = "Plasmid")

comp3$Plasmid_col <- fct_reorder(comp3$Plasmid_col,
                                 comp3$Plas_group)
comp3stat$Plasmid_col <- factor(comp3stat$Plasmid_col,
                                levels = levels(comp3$Plasmid_col))


pd <- position_jitterdodge(dodge.width=0.6, jitter.width = 0.1)
pd2<-position_dodge(width=0.6)

ggplot(comp3, aes(x=Plasmid_col,y=w_corr))+ 
  geom_point(position=pd2,size=4)+
  geom_hline(yintercept = 1, linetype="dashed")+
  theme_bw()+
  theme(axis.text=element_text(size=20),axis.title=element_text(size=25), legend.text=element_text(size=25),legend.title=element_text(size=25),
        axis.text.x = ggtext::element_markdown(angle = 45, hjust = 1))+labs(x="Plasmid", y="Relative Fitness (W)")+ 
  geom_pointrange(data=comp3stat,aes(x=Plasmid_col,ymin=w_corr-std,ymax=w_corr+std),size=2,alpha=0.3,position=pd2) +
   scale_y_continuous(breaks = seq(0.5, max(comp3$w_corr) + 0.2, by = 0.2))
  
ggsave("Comp_fit.png",dpi=1200,width=12,height=10)
```

Analysing fitness values

```{r}
fit <- comp3

F3 <- aov(w_corr ~ Plasmid, data = fit)

summary(F3)
TukeyHSD(F3)
```

---

**[Back to index.](../4_Analysis.md)**
